Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks

### 2019-04-18 doc/paper/paper.tex
Updates from Overleaf

\newcommand{\echonn}{\texttt{echonn}}
\newcommand{\der}[2][t]{\frac{\mathrm{d}#2}{\mathrm{d}#1}}

@@

An example of this is the trajectories of the sun and planets in our solar
system. Newton's equations can be used to create a system of equations to
model the bodies of mass (known as the n-body problem). However, with the
existence of other bodies of mass from other solar systems within the
universe, small perturbations are introduced into the system. If the system
is chaotic, then over time, the actual trajectories will diverge from our
model's \cite{oestreicher2007history}.
One of the defining characteristics of chaos is that the error from
perturbations grows exponentially \cite{oestreicher2007history}. At the time
of Lorenz's paper, unstable error was not well understood, and this led to
Lorenz's surprising discovery \cite{oestreicher2007history}. Lorenz reported
that a single iteration in calculating the solution to the dynamical system
took approximately one second \cite{lorenz1963deterministic}. In his paper,
he shows one of his calculations having about 3000 iterations, which must
have taken around 50 minutes. Given the amount of time it takes to perform
the calculation, it's tempting to continue the calculation from the output of
the previous calculation. However, the story of the discover is that while
the computer used 6 digit accuracy, it only output 3 digits, so when the
simulation was continued by Lorenz with the less accurate measurements, he
found the results to be very different \cite{oestreicher2007history}.

@@

In terms of forecasting, it's useful to know how chaotic a dynamical system
is. Chaos will be measured by the exponential rate at which nearby
trajectories diverge. Constants describing the divergence are known as the
Lyapunov Characteristic Exponents (LCEs) \cite{sandri1996numerical}. There is
an LCE for each dimension, and since they all represent exponential growth.
The error growing at the rate of the largest LCE will dominate over the
others.

Given a small perturbation $\epsilon$ to a dynamical system in the direction
of the largest LCE, we can approximate the growth of the perturbation by the
equation $\epsilon(\Delta t) = \epsilon_0 e^{\lambda \Delta t}$ where
$\lambda$ is the largest LCE \cite{bezruchko2010extracting}. For an LCE of
zero, then the error doesn't grow exponentially, and thus, there is no chaos.
Error created from smaller LCEs is negligible. The approximate time scale
where perturbations become large is going to be proportional to $1 /
\lambda$. This is known as the Lyapunov time \cite{bezruchko2010extracting}.
For our solar system, the Lyapunov time is 10,000,000 years
\cite{oestreicher2007history}. Researchers have used it to evaluate the
effectiveness of forecasting models on chaotic problems
\cite{pathak2018model}. Predicting the solution of a chaotic model over an
interval of time which is multiple times the size of the Lyapunov time shows
the predictability of a model.

@@

In order to find the Lyapunov time, we only need to solve for the largest
LCE. To find this, we will use the following method detailed in
\cite{viswanath1998lyapunov}.

@@

where $\Mv(t) \in \mathbb{R}^{d \times d}$ and $\frac{\partial \fv}{\partial
\xv}$ is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$.

@@

Implemented in \echonn, the algorithm was tested on the Lorenz system for
various parameters and initial conditions. The LCEs match the expected values
very accurately (see the Lorenz System section: \ref{sec:lorenz}). The 3
dimensional 3-body problem is an 18 dimensional dynamical system, so while
the dynamical system is implemented in \echonn, its max LCE can't be
calculated accurately since the resultant system has 342 dimensions. The
algorithm was also tested on the differential equation which produces a
circle:

@@

Empirically shown, \echonn can be used to identify the max LCE of smaller
dynamical systems, and special precaution has to be taken when choosing a
time interval. Using a larger time will yield a more accurate $\lambda_1$,
but it may also cause an overflow if the system is chaotic. Thus, maximum
time values need to be chosen such that $\vert \vert \Mv\vert \vert_{\infty}$
doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert \vert_{\infty}$ is
the $\ell^\infty$ norm).

@@

In order to generate the data to be analyized, the solutions to these
dynamical systems need to be computed. Each of the systems are defined by a
system of differential equations which can be numerically solved. However,
due to the chaotic nature, error introduced by the numerical solution will
grow exponentially, so precaution need to be taken to avoid having this sort
of error introduced into the collected data. When numerically solving chaotic
problems, it's possible for the chaotic behavior to be suppressed, or where
there shouldn't be chaotic behavior, it could be created
\cite{corless1994good}.

@@

Not all parameters for the Lorenz equation will produce chaotic behavior. In
Lorenz's article, he uses the parameter values $\sigma=10$, $\beta=8/3$, and
$\rho=28$ \cite{lorenz1963deterministic}, and it has a maximum LCE of
$0.90566$ \cite{viswanath1998lyapunov}. There are other parameters which
produce chaotic results outlined in Table \ref{table:lorenz_params}. Also
shown in Table \ref{table:lorenz_params} are the values calculated by my
Python module \echonn are listed. These values are not as accurate, but
demonstrate the sufficient accuracy of \echonn.

@@

@@

The 3-body problem, or the generalized n-body problem, is a famous problem
which many have analyzed; most notably, Herni Poincaré wrote many papers
developing the area significantly \cite{chenciner2000remarkable}. One of
Poincaré's solutions, known as the circular restricted 3-body problem,
demonstrates chaotic properties \cite{oestreicher2007history}. Poincaré
performs several reductions of the original 3-body problem in order to come
to a much more useful form of the problem. The mass of one of the objects is
assumed to be negligible. This is the case for Sun-Earth-Moon orbits. The
moon has little effect on the orbit of the sun. Another aspect is the
rotating reference frame. Since the mass of one of the bodies is negligible,
the problem simplifies to first solving a two body problem. While much easier
to solver, Poincaré went further to stop the motion of the two planets by
using a rotating reference frame. Rotating with the planets, the problem can
be modeled seemingly by two static planets and a third mass-less planet
orbiting around those in a two dimensional plane.

@@

There are however two issues when integrating this problem. As objects
approach each other, the distance between them approaches zero. If collision
occurs, then this approaches infinity. There is no collision detection in
this model, and each body has a radius of zero, so this case is not handled.
Instead, a divide by zero error occurs, and the model cannot be calculated
further.

@@

While it is possible to use a symplectic integrator, it is not necessary.
Under the right initial conditions, the third body can diverge regardless
(fast initial velocity away from the bodies). The simple divergent spiral
isn't chaotic, so instead orbital and collision less initial conditions will
be used. Without divergence and without collisions, the symplectic integrator
is unnecessary.

@@

Although, it's rare that the output units are calculated from previous output
units \cite{jaeger2002tutorial}. Sometimes for training, the output units are
set to the value of $dv$, called teacher forcing, so that the model trains on
actual values as opposed to its accumulated errors \cite{jaeger2002tutorial}.

@@

How the units are connected can vary, but the suggested way
\cite{jaeger2002tutorial} is to connect the internal layer to the current
input layer, the previous internal layer, and the previous output layer.
These connections will have weights described in \ref{sec:rnn}. Respectively,
these weights will be denoted by the matrices: $\Win$, $\W$, and $\Wback$.
The activation function for the internal layer is $\tanh$.

The output layer is connected to the current input layer and current internal
layer. Additionally, a bias input is connected to this layer. The weight for
these connections will be denoted $\Wout$ (we only have one matrix since the
connected layers are concatenated into a single vector, per Jaeger's notation
\cite{jaeger2002tutorial}). The activation function for these connections is
linear. Although, the recommended choice is $\tanh$, the output layer doesn't
necessarily lie in the range of $\tanh$, so it can't be used without some
form of normalization.

@@

@@

This suggests that as the model iterates, the internal state converges to
zero.

@@

Second, the initialization data is fed into the model. This initializes the
inner state of the model, and the initial state (zeros) is lost due to the
echo state property. $T_0$ should be chosen such that this is truly the case.
Like $\alpha$, $T_0$ is a parameter of the echo state network and differs
depending on the problem. Good values for $T_0$ are 10 to 500 depending on
the problem complexity.

Third, the training values are fed into the model to sample the network. We
can store the input unit values, the internal state values, and the desired
output values as matrices $M$ and $T$. $M$ will store the input unit values
and internal state value pairs $(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots,
(\uv(t_T), 1, \xv(t_T))$ in its rows. $T$ will store the output unit values
$\dv(t_{T_0}), \dots, \dv(t_T)$.

To train the model, we want $\Wout M$ to approximate accurately $g^{-1}(T)$,
where $g$ is the activation function for the output layer. Since we use an
idempotent output layer activation function, $g^{-1}(T) = T$.

@@

First, data from the models is collected. Using the process described above
(Section \ref{sec:calc_lya}), the Largest Lyapunov Exponent is calculated,
$\lambda_1$. Then, the Lyapunov Time is calculated by $\frac{1}{\lambda_1}$.
We will then integrate the model with an algorithm called RK4(5) over the
interval $[0, \frac{50}{\lambda_1}]$ with time steps of $\frac{1}{100
\lambda_1}$. This will result in 5,000 data points.

@@

The first data point of the training set is $t_0 = 0$. The parameter $T_0$
marks where the model begins training on the data (see Section
\ref{sec:etrain}). Using slightly different notation to account for the extra
validation set, we will denote the first validation time point by $t = T$. We
will denote the first testing time point by $t = T_f$

@@

However, the $\text{RMSE}_\text{test}$ is calculated slightly different since
we want to know how far the model can predict before deviating too far. The
$\text{RMSE}_\text{test}(t_f)$ will be calculated on the interval $[T_f,
t_f)$ and the prediction uses teacher-forced values on $[0, T_f)$. Lastly,
$\text{RMSE}_\text{test}$ will notate the RMSE for the entire test set.

We will look at how far past $T_f$ the ESN model's $\text{RMSE}_\text{test}$
grows too large. The best model will be the model with the lowest Testing
RMSE for the longest time.

\subsection{Results}
\subsubsection{Lorenz}

To integrate the model, I use the initial conditions: $(x_0, y_0, z_0) =
(-13.78990503, -11.6069343, 36.17332518)$. The calculated Lyapunov value was
$0.9348566$.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r0_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r0_rmse}
    \end{subfigure}
    \caption{The model diverges when entering never seen region.}
    \label{fig:lorenz_r0}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r1_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r1_rmse}
    \end{subfigure}
    \caption{The model continues in wrong direction after intersection.}
    \label{fig:lorenz_r1}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r2_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r2_rmse}
    \end{subfigure}
    \caption{The model has completely deviated after an intersection.}
    \label{fig:lorenz_r2}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r3_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r3_rmse}
    \end{subfigure}
    \caption{Fits model really well, but follows wrong wings.}
    \label{fig:lorenz_r3}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r4_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r4_rmse}
    \end{subfigure}
    \caption{Fits model really well, but follows wrong wings.}
    \label{fig:lorenz_r4}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_6_param_222_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r6_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_6_param_222_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r6_rmse}
    \end{subfigure}
    \caption{Doesn't fit model smoothy and eventually diverges.}
    \label{fig:lorenz_r6}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_7_param_73_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r7_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_7_param_73_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r7_rmse}
    \end{subfigure}
    \caption{Sporadic behavior shown long after divergence.}
    \label{fig:lorenz_r7}
\end{figure}

@@

The double pendulum's initial condition was

\begin{equation}
(\theta_1, \theta_2, \omega_1, \omega_2) 
= (4.66648088, 4.29820501,
8.01585663 \times 10^{-4}, 3.71175422 \times 10^{-3}).
\end{equation}

In this case, we need $\theta_1$ large enough that the system is chaotic, but
the rest of the parameters were chosen randomly (with rotational velocities
close to zero). The resulting Lyapunov Exponent is $0.60074$. Like the
Lorenz, the differences between time points were used to train the ESN since
the ESN was showing poor results otherwise. As the pendulum kept looping the
center, $\theta_1$ and $\theta_2$ kept growing; this shouldn't be an issue,
but using the differences keeps the system variables in a closer range.

Figure \ref{fig:doub_pend_diff} shows a plot of each of the dataset's
dimensions. It seems like the dataset's activity is fairly sparse. There is
only gradual but seemingly random variations until a few spots around $t=10,
30, 60, \text{and} 75$. The results for the ESN weren't very strong, and none 
could predict more than one Lyapunov Times before diverging.

Because there are portions of the dataset which are calm, the ESN then has to
pickup on signals which lead to the spikes in movement. We see the testing
graph that while it predicts the testing set well, it eventually tries to
predict spikes but does so at the wrong time, and the spike pattern never
ends, has too large of magnitude, and is very rapid. Overall the model
performs poorly, and it's reflected in the RMSE. The likely reason for this
is that the ESN can't learn the complex patterns from such a small and sparse
dataset.

@@

Generating data for the 3 body problem was more difficult since we have to becareful that the trajectory doesn't diverge or collide with the other bodies.

>>>>

\newcommand{\echonn}{\texttt{echonn}}
\newcommand{\echonnsp}{\texttt{echonn}\text{ }}
\newcommand{\der}[2][t]{\frac{\mathrm{d}#2}{\mathrm{d}#1}}

@@

An example of this is the trajectories of the sun and planets in our solar
system. Newton's equations can be used to create a system of equations to
model the bodies of mass (known as the n-body problem). However, with the
existence of other bodies of mass from other solar systems within the
universe, small perturbations are introduced into the system. If the system
is chaotic, then over time, the actual trajectories will diverge from our
model's \cite{oestreicher2007history}.
One of the defining characteristics of chaos is that the error from
perturbations grows exponentially \cite{oestreicher2007history}. At the time
of Lorenz's paper, unstable error was not well understood, and this led to
Lorenz's surprising discovery \cite{oestreicher2007history}. Lorenz reported
that a single iteration in calculating the solution to the dynamical system
took approximately one second \cite{lorenz1963deterministic}. In his paper,
he shows one of his calculations having about 3000 iterations, which must
have taken around 50 minutes. Given the amount of time it takes to perform
the calculation, it's tempting to continue the calculation from the output of
the previous calculation. However, the story of the discover is that while
the computer used 6 digit accuracy, it only output 3 digits, so when the
simulation was continued with the less accurate measurements, Lorenz
found the results to be very different \cite{oestreicher2007history}.

@@

It's useful to know how chaotic a dynamical system
is. Chaos is usually measured by the exponential rate at which nearby
trajectories diverge. Constants describing the divergence are known as the
Lyapunov Characteristic Exponents (LCEs) \cite{sandri1996numerical}.
There is
an LCE for each dimension, and since they all represent exponential growth,
the error growing at the rate of the largest LCE will dominate over the 
others.
Given a small perturbation $\epsilon$ to a dynamical system in the direction
of the largest LCE, we can approximate the growth of the perturbation by the
equation $\epsilon(\Delta t) = \epsilon_0 e^{\lambda \Delta t}$ where
$\lambda$ is the largest LCE \cite{bezruchko2010extracting}. For an LCE of
zero, then the error doesn't grow exponentially, and thus, there is no chaos.
Error created from smaller LCEs is negligible. The approximate time scale
where perturbations become large is going to be proportional to $1 /
\lambda$. This is known as the Lyapunov time \cite{bezruchko2010extracting}.
For example, the Lyapunov time of the orbits of our solar system is
10,000,000 years
\cite{oestreicher2007history}.

Researchers have used it to evaluate the
effectiveness of forecasting models on chaotic problems in the past.
Predicting the solution of a chaotic model over an
interval of time which is multiple times the size of the Lyapunov time shows
the predictability of a model. One group of researchers studied the 
Kuramoto-Sivashinsky chaotic model near an attractor and found that with a max
LCE of 0.1, they were able to predict 8 Lyapunov times before their forecast 
diverged \cite{pathak2018model}.

@@

In order to find the Lyapunov time, we only need to solve for the largest
LCE. To find this, we will use the following method
\cite{viswanath1998lyapunov}.

@@

where $\Mv(t) \in \mathbb{R}^{d \times d}$ and
$\frac{\partial \fv}{\partial \xv}$
is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$ and with 
$\Mv(0)$ equal to the identity matrix.

@@

Implemented in \echonn, the algorithm was tested on the Lorenz system for
various parameters and initial conditions. The LCEs match the expected values
very accurately (see the Lorenz System section: \ref{sec:lorenz}). The 3
dimensional 3-body problem is an 18 dimensional dynamical system, so while
the dynamical system is implemented in \echonn, its max LCE can't be
calculated accurately since the resulting system has 342 dimensions. The
algorithm was also tested on the differential equation which produces a
circle:

@@

Empirically shown, \echonnsp can be used to identify the max LCE of smaller
dynamical systems, and special precaution has to be taken when choosing a
time interval. Using a larger time will yield a more accurate $\lambda_1$,
but it may also cause an overflow if the system is chaotic. Thus, maximum
time values need to be chosen such that $\vert \vert \Mv\vert \vert_{\infty}$
doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert \vert_{\infty}$ is
the $\ell^\infty$ norm).

@@

In order to generate the data to be analyized, the solutions to these
dynamical systems need to be computed. Each of the systems are defined by a
system of differential equations which can be numerically solved. However,
due to the chaotic nature, error introduced by the numerical solution will
grow exponentially, so precaution needs to be taken to avoid having this sort
of error introduced into the collected data. When numerically solving chaotic
problems, it's possible for the chaotic behavior to be suppressed, or where
there shouldn't be chaotic behavior, it could be created
\cite{corless1994good}.

@@

Not all parameters for the Lorenz equation will produce chaotic behavior. In
Lorenz's article, he uses the parameter values $\sigma=10$, $\beta=8/3$, and
$\rho=28$ \cite{lorenz1963deterministic}, and it has a maximum LCE of
$0.90566$ \cite{viswanath1998lyapunov}. There are other parameters which
produce chaotic results outlined in Table \ref{table:lorenz_params}. Also
shown in Table \ref{table:lorenz_params} are the values calculated by my
Python module \echonnsp are listed. These values 
demonstrate the accuracy of \echonn.

@@

% TODO: Graph of converging LCE

@@

The 3-body problem, or the generalized n-body problem, is a famous problem
which many have analyzed; most notably, Herni Poincaré wrote many papers
developing the area significantly \cite{chenciner2000remarkable}. One of
Poincaré's solutions, known as the circular restricted 3-body problem,
demonstrates chaotic properties \cite{oestreicher2007history}. Poincaré
performs several reductions of the original 3-body problem in order to come
to a much more useful form of the problem. The mass of one of the objects is
assumed to be negligible. This is the case for Sun-Earth-Moon orbits: the
moon has little effect on the orbit of the sun. Another aspect is the
rotating reference frame. Since the mass of one of the bodies is negligible,
the problem simplifies to first solving a two body problem. While much easier
to solver, Poincaré went further to stop the motion of the two planets by
using a rotating reference frame. Rotating with the planets, the problem can
be modeled seemingly by two static planets and a third mass-less planet
orbiting around those in a two dimensional plane.

@@

There are however two issues when integrating this problem. As objects
approach each other, the distance between them approaches zero. If collision
occurs, then the diverging trajectory approaches infinity. There is
no collision detection in
this model, and each body has a radius of zero, so this case is not handled.
Instead, a divide by zero error occurs, and the model cannot be calculated
further.

@@

While it is possible to use a symplectic integrator, it is not necessary.
Under the right initial conditions, the third body can diverge regardless
(e.g., fast initial velocity away from the bodies).
The simple divergent spiral
isn't chaotic, so instead orbital and collision less initial conditions will
be used. Without divergence and without collisions, the symplectic integrator
is unnecessary.

@@

Although, it's rare for the ESN that the output units are calculated
directly from previous output
units \cite{jaeger2002tutorial}. Sometimes for training, the output units are
set to the value of $\dv$, called teacher forcing, so that the model trains on
actual values as opposed to its accumulated errors \cite{jaeger2002tutorial}.

@@

How the units are connected can vary, but the suggested way
\cite{jaeger2002tutorial} is to connect the current
input layer, the previous internal layer, and the previous output layer to
the internal layer.
These connections will have weights described in \ref{sec:rnn}. Respectively,
these weights will be denoted by the matrices: $\Win$, $\W$, and $\Wback$.
The activation function for the internal layer is $\tanh$.

Connected to the output layer is the current input layer and current internal
layer. Additionally, a bias input is connected to this layer. The weight for
these connections will be denoted $\Wout$ (we only have one matrix since the
connected layers are concatenated into a single vector, per Jaeger's notation
\cite{jaeger2002tutorial}). The activation function for these connections is
linear. Although, the recommended choice is $\tanh$, the output layer doesn't
necessarily lie in the range of $\tanh$, so it can't be used without some
form of normalization.

The dimensions of the weight matrices are as follows:
$\Win \in \mathbb{R}^{N \times K}$,
$\W \in \mathbb{R}^{N \times N}$,
$\Wback \in \mathbb{R}^{N \times L}$, and
$\Wout \in \mathbb{R}^{N \times (K+1+N)}$,

@@

"Feeding data" into the model means iterating using values for $\uv$. If the 
feeding is teacher-forced, then the output layer's values are set to $\dv$. 
Otherwise, the calculated output values are used.

@@

This suggests that as the model iterates, the internal state converges to
zero. Also, $\tanh$ is contractive further supporting this idea.

@@

Second, the initialization data is fed into the model with teacher-forced 
output. This initializes the
inner state of the model, and the initial state (zeros) is lost due to the
echo state property. $T_0$ should be chosen such that this is truly the case.
Like $\alpha$, $T_0$ is a parameter of the echo state network and differs
depending on the problem. Good values for $T_0$ are 10 to 500 depending on
the problem complexity.

Third, the training values are fed into the model to sample the network with 
teacher-forced output. We define a matrix $M$ to store the input unit values,
bias, and internal state value pairs, $(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots,
(\uv(t_T), 1, \xv(t_T))$, in its rows. We define a matrix $T$ to store the 
output unit values: $\dv(t_{T_0}), \dots, \dv(t_T)$.

To train the model, we want $g(\Wout M)$ to approximate accurately 
$T$, where $g$ is the activation function for the output layer.
Since we use an
idempotent output layer activation function, $g^{-1}(X) = g(X) = X$, so

\begin{align*}
    g(\Wout M) &= T \\
    \Wout M &= g^{-1}(T) \\
    \Wout M &= T
\end{align*}

@@

First, data from the models is collected. Using the process described above
(Section \ref{sec:calc_lya}), the Largest Lyapunov Exponent is calculated,
$\lambda_1$. Then, the Lyapunov Time is calculated by $\frac{1}{\lambda_1}$.
We will then integrate the model with the algorithm RK4(5) over the
interval $[0, \frac{50}{\lambda_1}]$ with time steps of $\frac{1}{100
\lambda_1}$. This will result in 5,000 data points.

@@

The first data point of the training set is $t = 0$. The parameter $T_0$
marks where the model begins training on the data (see Section
\ref{sec:etrain}). Using slightly different notation to account for the extra
validation set, we will denote the first validation time point by $t = T$. We
will denote the first testing time point by $t = T_f$

@@

However, the $\text{RMSE}_\text{test}$ is calculated slightly different since
we want to know how far the model can predict before deviating too far. The
$\text{RMSE}_\text{test}(t)$ will be calculated on the interval $[T_f,
t)$ and the prediction uses teacher-forced values on $[0, T_f)$. Lastly,
$\text{RMSE}_\text{test}$ as a value instead of a function
will notate the RMSE for the entire test set.

We will look at how far past $T_f$ the ESN model's $\text{RMSE}_\text{test}$
grows too large. The best model will be the model with the lowest validation 
RMSE as well as a low testing RMSE.

\subsection{Results}

Overall, the systems proved too difficult to forecast; however, this could 
be due to lacking data. While more data can generated, numerical errors might 
grow too large. Since we have about 5,000 data points, we can't have an 
internal state dimension of more than 500 or we risk over-fitting 
\cite{jaeger2002tutorial}.

The Lorenz model showed strong forecasting results, but due to the shape of 
the attractor, it was easy for the small perturbations to cause quick 
divergence as the trajectories followed the incorrect "wing" of the butterfly.

The double pendulum model showed very poor results. Occasionally, the pendulum
would move much quicker causing spikes in the data. The ESN tried to predict 
when these spikes occurred, but was unable. Even the forecasted shape of the 
rapid movement was inaccurate. This is likely due to the lack of data.

The restricted circular three body problem performed much better than the
other two, but this is due to the existence of a simple attractor: a single 
circular orbit as opposed to two rings in the Lorenz system. Also, the max LCE
was much smaller for this system (0.1759 vs 0.9 and 0.6) suggesting that it 
was less chaotic. While
it's possible that the chaotic behavior was created by the numerical
solution, it suggests the strong predictability of lesser chaotic systems.

In the case of researchers performing a similar analysis mentioned before 
\cite{pathak2018model}, they too had strong performance for a less chaotic 
problem. I wasn't able to predict 8 Lyapunov times, but there might be 
reasons why my results weren't as strong other than the fact that
the ESN might need more data that I used.
The lorenz attractor is more chaotic than the attractor they used. The double 
pendulum lacks an attractor and shows a higher level of chaos. The 3 body
problem has a lower LCE and a simple attractor, and showed strong results as 
well.

\subsubsection{Lorenz}

To integrate the model, I use the initial conditions: $(x, y, z)_{t=0} =
(-13.78990503, -11.6069343, 36.17332518)$. The calculated Lyapunov value was
$\lambda_1 = 0.9348566$.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r0_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r0_rmse}
    \end{subfigure}
    \caption{Diverges when entering space outside training data.}
    \label{fig:lorenz_r0}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r1_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r1_rmse}
    \end{subfigure}
    \caption{Continues in wrong direction at intersection.}
    \label{fig:lorenz_r1}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r2_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r2_rmse}
    \end{subfigure}
    \caption{Deviates after an intersection.}
    \label{fig:lorenz_r2}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r3_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r3_rmse}
    \end{subfigure}
    \caption{Fits model well but follows wrong wings.}
    \label{fig:lorenz_r3}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r4_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r4_rmse}
    \end{subfigure}
    \caption{Fits model well but trajectory is ahead in time.}
    \label{fig:lorenz_r4}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_6_param_222_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r6_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_6_param_222_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r6_rmse}
    \end{subfigure}
    \caption{Diverges after intersection.}
    \label{fig:lorenz_r6}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_7_param_73_fit.png}
        \caption{The True Figure vs ESN Prediction}
        \label{fig:lorenz_r7_fit}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_7_param_73_rmse.png}
        \caption{RMSE Over Time}
        \label{fig:lorenz_r7_rmse}
    \end{subfigure}
    \caption{Sporadic behavior shown after divergence.}
    \label{fig:lorenz_r7}
\end{figure}

@@

In this case, we need $\theta_1$ large enough that the system is chaotic, but
the rest of the parameters were chosen randomly (with rotational velocities
close to zero). The double pendulum's initial condition was

\begin{equation}
(\theta_1, \theta_2, \omega_1, \omega_2)_{t=0} 
= (4.66648088, 4.29820501,
8.01585663 \times 10^{-4}, 3.71175422 \times 10^{-3}),
\end{equation}

and the resulting Lyapunov Exponent is $\lambda_1 = 0.60074$. Like the
Lorenz, the differences between time points were used to train the ESN since
the ESN was showing poor results otherwise. As the pendulum kept looping the
center, $\theta_1$ and $\theta_2$ kept growing; this shouldn't be an issue,
but using the differences keeps the system variables in a closer range.

Figure \ref{fig:doub_pend_diff} shows a plot of each of the dataset's
dimensions. It seems like the dataset's activity is fairly sparse. There is
only gradual but seemingly random variations until a few spots around
$t=10, 30, 60, \text{and } 75$. The results
for the ESN weren't very strong, and none 
could predict more than one Lyapunov Times before diverging.

Because there are portions of the dataset which are calm, the ESN then has to
pickup on signals which lead to the spikes in movement. We see in the 
graphs that while it predicts the testing set well, it eventually tries to
predict spikes but does so at the wrong time, and the spike pattern never
ends, has too large of magnitude, and is very rapid. Overall the model
performs poorly, and it's reflected in the RMSE which quickly diverges well above 3. The likely reason for this
is that the ESN can't learn the complex patterns from such a small and sparse
dataset. With more data or more complex internals, maybe the ESN could learn 
to forecast this dataset.

@@

Generating data for the 3 body problem is more difficult since we have to be 
careful that the trajectory doesn't diverge or collide with the other bodies.
